Learning and Evaluation of \strips\ Action Models with Classical Planning

### 2018-05-31 doc/aij18/background.tex
small updates to sections 1 and 2

This section serves two purposes. In subsection \ref{basic_planning} we introduce the basic planning concepts as well as the classical planning model that we will use throughout the rest of the paper. Section \ref{related_work} summarizes existing approaches to learn action models in classical planning and highlights the novelty of our contribution via the related work.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces the {\em state trajectory} $s=\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$ . A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.~if the goal condition is satisfied at the last state reached after following the application of the plan $\pi$ in the initial state $I$.

In this work, we will use the term \emph{observation} to refer to the input knowledge of a learning task. Hence, $s=\tup{s_0, s_1, \ldots, s_n}$ is a fully-observable (\FO) state trajectory if every state $s_i$ is a full assignment of values to fluents and the minimal action sequence to transit from state $s_i$ to state $s_{i+1}$ is composed of a single action; that is, $\theta(s_i,\tup{a})=s_{i+1}$. Otherwise, $s$ is a partially-observable (\PO) state trajectory, meaning that at least one $s_i$ is a partial assignment of values to fluents in which one or more literals are missing. This general definition gives rise to two particular cases:

\begin{itemize}
\item when all fluents of a state $s_i$ are missing, $s_i$ is a \emph{missing} or \emph{empty} state
\item when the $n-1$ intermediate states of a trajectory $s$ are missing, $s$ is a \emph{non-observable} (\NO) state trajectory
\end{itemize}

Hence, a partially-observed state trajectory can be interpreted as a sequence that contains $n+1$ possibly incomplete states induced by an $n$-action sequence, or as a sequence with some missing intermediate states, which can be regarded as a particular case of partial observability. We will refer to the former as a \PO trajectory and to the latter as a \POstar state sequence.

%In the rest of the paper, we will use the following notation to refer to the different levels of state visibility: $s$ denotes a \FO state trajectory; $s^{+}$ refers to a \PO state trajectory with $n+1$ possibly incomplete states; $s^{*}$ indicates a \PO state trajectory with $m$ missing intermediate states, $m < n-1$; and $s^{0}$ is a \NO state trajectory with $n-1$ missing states.

Regarding an action sequence, $\tup{a_1, \ldots, a_n}$ is a \FO action sequence if it contains all the necessary actions to achieve the goal from an initial state, a \PO action sequence if at least one necessary action to achieve the goal is missing in the sequence and a \NO action sequence in case the sequence is empty.

A \emph{plan trace} is a combination of a state and an action sequence $\tau = \langle s_0, a_1, s_1, a_2, s_2, \ldots, a_n, s_n \rangle$. A plan trace $\tau$ for a planning problem $P=\tup{F,A,I,G}$ satisfies $s_0=I$, $G \in s_n$ and every element in the action sequence of $\tau$ is an action of $A$. Additionally, plan traces will be classified accordingly to the type of state trajectory (\FO, \PO, \POstar or \NO) and action sequence (\FO, \PO or \NO).

%Additionally, $\tau$ is a plan trace with PO states if $\tup{s_0, s_1, \ldots, s_n}$ is a PO state sequence; and a plan trace with PO actions if $\tup{a_1, \ldots, a_n}$ is a PO action sequence.

>>>>

This section serves two purposes: First we introduce the basic planning concepts as well as the classical planning model that we will use throughout the rest of the paper. Second, we summarize the existing approaches to learn action models in classical planning and highlight our contributions via the related work.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\in\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that induces the {\em state trajectory} $s=\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$ . A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$, i.e.~if the goal condition is satisfied at the last state reached after following the application of the plan $\pi$ in the initial state $I$. A solution plan is {\em optimal} if it has minimum length.

In this work, we will use the term \emph{plan trace} to refer to the observation of a plan execution that starts on a given initial state. \emph{Plan traces} are the input knowledge of the learning tasks that are addressed in this paper. A \emph{plan trace} is a combination of a state and an action sequence $\tau = \langle s_0, a_1, s_1, a_2, s_2, \ldots, a_n, s_n \rangle$. This sequence is obtained watching the execution of a plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applied in state $s_{i-1}$ and generates the successor state $s_i$. We name different classes of \emph{plan traces} according to the kind of knowledge that is observed:
\begin{itemize}
  \item[{\bf\FO}] We say that a plan trace $\mathcal{T}=\tup{s_0, s_1, \ldots, s_n}$ is a {\em fully-observable} (\FO) {\em state trajectory} if every state $s_i$, {\small $1\leq i\leq n$}, is a full assignment of values to fluents and the transition from states $s_i$ to states $s_{i+1}$ is caused by the execution of a single action $a_i$ that is unobserved; that is, $\theta(s_i,\tup{a_i})=s_{i+1}$. 
  \item[{\bf\PO}] A plan trace $\mathcal{T}$ is a {\em partially-observable} (\PO) state trajectory, meaning that at least one $s_i\in\mathcal{T}$ is a partial assignment of values to fluents in which one or more literals are missing, formally there exists a state $s_i$, {\small $1\textless i$} such that $|s_i|\textless|F|$. 
  \item[{\bf\POstar}] \emph{Plan traces} with some missing intermediate states is regarded as a particular case of partial observability (that we name \POstar). 
  \item[{\bf\NO}]A state trajectory $\mathcal{T}$ where every intermediate state $s_i\in\mathcal{T}$, {\small $1\textless i$}, holds that $|s_i|=0$ is a \emph{non-observable} (\NO) state trajectory.
\end{itemize}

Regarding to the action sequence, $\tup{a_1, \ldots, a_n}$, it is called a \FO action sequence if it contains all the necessary actions $a_i$ to transit every state $s_i$ to the corresponding succesor state $s_{i+1}$, for each {\small $1\leq i\leq n$}. We say that an action sequence is \PO if at least one of these necessary $a_i$ actions is missing in the sequence and \NO, in case that the action sequence is empty.

### 2018-05-31 doc/aij18/introduction.tex
small updates to sections 1 and 2

\section{Introduction}
\label{sec:Section1}

Besides {\em plan synthesis}~\cite{ghallab2004automated}, planning action models are also useful for {\em plan/goal recognition}~\cite{ramirez2012plan}. At both planning tasks, an automated planner is required to reason about action models that correctly and completely capture the possible world transitions~\cite{geffner:book:2013}. Unfortunately, building planning action models is complex, even for planning experts, and this knowledge acquisition task is a bottleneck that limits the potential of AI planning~\cite{kambhampati:modellite:AAAI2007}.

Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of \strips\ action models, the vanilla action model for automated planning~\cite{fikes1971strips}, is not straightforward though:
\begin{itemize}
\item The {\em input} to ML algorithms (the {\em learning/training} data) usually are finite vectors encoding the value of fixed features in a given set of objects. The input for learning planning action models are observations of plan executions (where each plan possibly has a different length and may involve a different number of objects).
\item The {\em output} of ML algorithms usually is a scalar value (an integer, in the case of {\em classification} tasks, or a real value, in the case of {\em regression} tasks). When learning \strips\ action models the output is, for each action, the sets of {\em preconditions}, {\em negative} and {\em positive effects} that define which state transitions are possible.
\end{itemize}

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016generalized,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning \strips\ action models that can be defined as a classical planning compilation. A solution to the classical planning task that results from our compilation is a sequence of actions that determines the preconditions and effects of a \strips\ model such that this \strips\ model satisfies a set of plan executions that is given as input.

The compilation approach is appealing by itself because it leverages off-the-shelf planners for learning and because its practicality allow us to report learning results over a wide range of domains from the {\em International Planning Competition} (IPC). Moreover, it opens up a way towards \emph{bootstrapping} planning action models, enabling a planner to gradually learn/update its action model. Apart from these, our classical planning compilation presents the following contributions:
\begin{enumerate}
\item {\em Input flexibility}. Our classical planning compilation is flexible to various amount and kind of available input knowledge. The action model to learn can be partially specified and the learning examples can range from a set of plans (with their corresponding initial state) or state observations, to just a set of initial and final states where no intermediate action or state is observed. % Also accepts background knowledge

\item {\em Model validation}. The compilation poses a novel framework to assess the validation of a STRIPS model with respect a plan execution. This validation capacity goes beyond the functionality of VAL~\cite{howey2004val} since it does require neither a full action model nor a fully observed plan to determine validation.

\item {\em Model evaluation}. The compilation can assess how well a given \strips\ action model matches a plan execution, which allows us to assess learning performance without having the actual action model. The idea is to assess the amount of edition required by the input action model to induce the given observation of the plan execution. 
\end{enumerate}

A first description of the compilation previously appeared in the conference paper~\cite{aineto2018learning}. Compared to that conference paper, this work includes the following novel material:
\begin{itemize}
\item A unified formulation for learning and evaluating \strips\ action models from observed executions of plans. 
\item Redefining the {\em precision} and {\em recall} metrics to evaluate learned action models with respect to a plan execution.
%\item Leveraging diverse forms of {\em background knowledge} for learning and evaluating \strips\ action models.
\item A complete empirical evaluation of the compilation approach. Our evaluation analyses how the amount of input knowledge affects to the performance of the compilation approach when learning and evaluating \strips\ action models.
\end{itemize}

Section~\ref{sec:Section2} reviews related work on learning planning action models. Section~\ref{sec:Section3} introduces the classical planning model and the \strips\ action model (the target of our learning and evaluation tasks). Section~\ref{sec:Section4} formalizes the learning of \strips\ action models with regard to different amount and kind of available input knowledge. Sections~\ref{sec:Section5} and ~\ref{sec:Section6} describe our compilation approach for addressing the formalized learning tasks its extension to evaluate and recognize \strips\ action models. Section~\ref{sec:Section7} reports the data collected in a two-fold empirical evaluation of our learning approach: First, the learned \strips\ action models are tested with observations of plan executions and second, the learned models are compared to the actual models. Finally, Section ~\ref{sec:Section8} discusses the strengths and weaknesses of the compilation approach and proposes several opportunities for future research.

>>>>

\section{Introduction}
\label{sec:introduction}

Besides {\em plan synthesis}~\cite{ghallab2004automated}, planning action models are also useful for {\em plan/goal recognition}~\cite{ramirez2012plan}. At both planning tasks, automated planners reason about action models that correctly and completely capture the possible world transitions~\cite{geffner:book:2013}. Unfortunately, modeling planning actions is complex, even for planning experts, and this knowledge acquisition task is a bottleneck that limits the potential of AI planning~\cite{kambhampati:modellite:AAAI2007}.

Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of \strips\ action models, the vanilla action model for automated planning~\cite{fikes1971strips}, is not straightforward though:
\begin{itemize}
\item The {\em input} to ML algorithms (the {\em learning/training} data) usually are finite vectors encoding the value of fixed features in a given set of objects. The input for learning planning action models are observations of plan executions (where each observed plan possibly has a different number of steps and involves a different number of objects).
\item The {\em output} of ML algorithms usually is a scalar value (an integer, in the case of {\em classification} tasks, or a real value, in the case of {\em regression} tasks). When learning \strips\ action models the output is, for each action, the sets of {\em preconditions}, {\em negative} and {\em positive effects} that define the possible state transitions.
\end{itemize}

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016generalized,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning \strips\ action models that can be defined as a classical planning compilation. A solution to the classical planning task that results from our compilation is a sequence of actions that determines the preconditions and effects of a \strips\ model such that this model satisfies the plan executions given as input.

The compilation approach is appealing by itself because it leverages off-the-shelf planners and because its practicality allow us to report learning results over a wide range of domains from the {\em International Planning Competition} (IPC). Moreover, it opens up a way towards \emph{bootstrapping} planning action models, enabling a planner to gradually learn/update its action model. Apart from these, our compilation exhibits the following contributions:
\begin{enumerate}
\item {\bf\em Input flexibility}. Our classical planning compilation is flexible to various amount and kind of available input knowledge. The action model to learn can be partially specified and the learning examples can range from a set of plans (with their corresponding initial state) or state observations, to just a set of initial and final states where no intermediate action or state is observed. % Also accepts background knowledge

\item {\bf\em Model validation}. The compilation poses a novel framework to assess the validation of a STRIPS model with respect a plan execution. This validation capacity goes beyond the functionality of VAL~\cite{howey2004val} since it does require neither a full action model nor a fully observed plan to determine validation.

\item {\bf\em Model evaluation}. The compilation can assess how well a given \strips\ action model matches a plan execution, which allows us to assess learning performance without having the actual action model. The idea is to assess the amount of edition that is required by the input action model to induce the given observation of the plan execution. 
\end{enumerate}

A first description of the compilation previously appeared in our conference paper~\cite{aineto2018learning}. Compared to that conference paper, this work includes the following novel material:
\begin{itemize}
\item A unified formulation for learning and evaluating \strips\ action models from observed executions of plans. Further, these executions can only comprise state observations.
\item A redefinition of the ML metrics {\em precision} and {\em recall} to evaluate \strips\ action models with respect to observations of plan executions.
%\item Leveraging diverse forms of {\em background knowledge} for learning and evaluating \strips\ action models.
\item A complete empirical evaluation of the compilation approach. Our evaluation analyses how the amount of input knowledge affects to the performance of the compilation approach when learning and evaluating \strips\ action models. 
\end{itemize}

Section~\ref{sec:background} introduces classical planning and reviews related work on learning planning action models. Section~\ref{sec:Section3} introduces the classical planning model and the \strips\ action model (the target of our learning and evaluation tasks). Section~\ref{sec:Section4} formalizes the learning of \strips\ action models with regard to different amount and kind of available input knowledge. Sections~\ref{sec:Section5} and ~\ref{sec:Section6} describe our compilation approach for addressing the formalized learning tasks its extension to evaluate and recognize \strips\ action models. Section~\ref{sec:Section7} reports the data collected in a two-fold empirical evaluation of our learning approach: First, the learned \strips\ action models are tested with observations of plan executions and second, the learned models are compared to the actual models. Finally, Section ~\ref{sec:Section8} discusses the strengths and weaknesses of the compilation approach and proposes several opportunities for future research.
